18.821 - Poker Chips Finite board configurations

### 2014-09-19 FiniteBoards.tex
Added small earthquake
Also attempted to provide more structure/ordering of definitions.

In this file, we analyze the case where the initial number of chips is finite. That is, the sum of all the chips on the board is finite. 
In this file, we analyze the case where the initial number of chips is finite.
 That is, the sum of all the chips on the board is finite.

\begin{definition}
We say that a board is stable if its values are strictly less than 4. This is equivalent to earthquakes having no effect on the board. 
\end{definition}

\begin{theorem}
\label{finitestability}
If the initial board has a finite number of chips, there exists an $N$ such that after $N$ earthquakes, the resultant board will be in a stable.
\end{theorem}

The proof will be as follows: first, we will show that finite boards cannot return to a previous state. Let $B$ be the set of all possible board configurations, that is, $B = \{ f: \mathbb{Z} \times \mathbb{Z} \rightarrow \mathbb{Z}^{\geq 0}\}$. Each board configuration is given by a function of the integer lattice to the nonnegative integers. So $f(x,y)$ is the number of chips in the grid location $x,y$.

The earthquake is defined as $E: B \rightarrow B$ acts in the following manner:
\[ E(f(x,y)) = f(x,y) + \sum_{l \in n(x,y)} [[ f(l) \geq 4 ]] - 4[[f(x,y) \geq 4]] \]
Where $n(x,y)$ is the set of neighbors of $x,y$, and $[[ X ]] = 1$ if $X$ happened and $0$ otherwise. Let $B_f \subset B$ with all boards with a finite number of chips. 
\[ B_f = \bigcup_{n \geq 0} \{ f \in B | \sum_{x,y} f(x,y) = n \} \]

So a restatement of the theorem is to say that for all $f \in B$ where $\sum_{x,y} f(x,y) = S$, there exists an $N$ such that for all $m \in \mathbb{Z}$, $E^N(f) = E^{N+m}(f)$.

\begin{lemma}
If $E(f) \neq f$, then $E^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
Let $\Phi: B_f \rightarrow \mathbb{Z}$ be the following function
\[ \Phi(f) = \sum_{x,y} f(x,y)(x+y)^2 \]
Note that this is well-defined since the boards contain a finite number of chips. We will show that 
\[ E(f) \neq f \Rightarrow \Phi(E(f)) > \Phi(f) \]
This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.

We can evaluate $\Phi(E(f)) - \Phi(f)$.
\[ \Phi(E(f)) = \sum_{x,y} E(f(x,y)) (x+y)^2 \]
\[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (x+y)^2 \]
Suppose location $f(x,y) \geq 4$, then the number of chips in location $(x,y)$ will have decreased after applying $E$. For each decrease in the number of chips atlocation $(x,y)$ of $4$, there is an increase in its four neighbors neighbors by $1$.

Let $A$ be the set of active locations, or locations which $f(x,y) \geq 4$, which is non-empty by assumption. 
So
\[ \Phi(E(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(x+y)^2 + 2(x+y+1)^2 + 2(x+y-1)^2 \]
\[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
									   +2x^2 & +2y^2 & +4xy & +2x& +2y &+ 2\\
									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} \]
\[ \geq 4 > 0 \] 
\end{proof}

\begin{lemma}
\label{finiteextension}
Suppose $\sum_{x,y} f(x,y) = S$, and all chips are within the range $[x_l, x_h] \times [y_l, y_h]$. Then $E^n(f)$ only contains chips within the range $[x_l - \log S, x_h + \log S] \times [y_l-\log S, y_h+\log S]$ for all $n$.
\end{lemma}

@@

\begin{proof}
(of Theorem~\ref{finitestability}) Suppose $f \in B$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$. Then by Lemma~\ref{finiteextension}, we have that all $E^n(f)$ will lie in $R = [x_l - \log S, x_h + \log S] \times [y_l - \log S, y_h + \log S]$.

There exists less than $(x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ possible boards $b \in B$ that satisfy this. We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
\end{proof}

>>>>

In this file, we analyze the case where the initial number of chips is finite.
That is, the sum of all the chips on the board is finite.

\section{Preliminaries}
\begin{definition} A board is a map from the integer lattice to the non-negative integers.
That is a Board $f$ is an element of $\mathcal{B} := \text{Hom}(\mathbb{Z} \times \mathbb{Z}, \mathbb{Z}_{\geq 0})$.
The input values of a board represent the coordinates on the integer lattice that the lower left hand corner of square occupies and the output value refers to the number of chips on that square. 
A square is an element of $\mathbb{Z} \times \mathbb{Z}$. 
\end{definition}

\begin{definition} We define the big earthquake operator as a map from the set of all boards to the set of all boards and write $BE: \mathcal{B} \rightarrow \mathcal{B}$. 
It transforms all of the squares on the board with the following rule.
\begin{equation}
BE(f(x,y)) = f(x,y) - 4[[f(x,y) \geq 4]] + [[f(x-1,y) \geq 4]] + [[f(x+1,y) \geq 4]] + [[f(x,y-1) \geq 4]] + [[f(x,y+1) \geq 4]],
\end{equation}
where $[[A]]$ is defined as 
\begin{equation}
[[A]] = 
\begin{cases} 
1 \text{ if } A \text{ is true} \\ 
0 \text{ if } A  \text{ is false}
 \end{cases}.
\end{equation}
\end{definition}

\begin{definition}
 We define the small earthquake operator as a map from the set of all boards to the set of all boards and write $SE: \mathcal{B} \rightarrow \mathcal{B}$.
It selects one square $(x,y)$ with $f(x,y) \geq 4$ and transforms it and all of its neighbors with the following rule: 
\begin{align*}
SE( f(x,y)) = f(x,y)-4 \\
SE( f(x+1,y)) = f(x+1,y)+1 \\
SE( f(x-1,y)) = f(x-1,y)+1 \\
SE( f(x,y+1)) = f(x,y+1)+1 \\
SE( f(x,y-1)) = f(x,y-1)+1,
\end{align*}
and leaves all other squares unchanged.
\end{definition}

\begin{definition}
We say that a board is stable if its values are strictly less than 4. This is equivalent to both of the earthquake maps having no effect on the board. 
\end{definition}

\begin{definition} 
We say that a board is finite if the total number of chips on it is finite. That is a board $f$ is finite if the sum
\begin{equation}
S= \sum_{(x,y) \in \mathbb{Z} \times \mathbb{Z}} f(x,y) 
\end{equation}
is finite. 
We denote the set of finite boards as $\mathcal{B}_f$.
\end{definition}

\begin{definition}
We define the function $n: \mathbb{Z} \times \mathbb{Z} \rightarrow \mathbb{Z}^2 \times \mathbb{Z}^2 \times \mathbb{Z}^2 \times \mathbb{Z}^2 $ as 
\begin{equation}
n(x,y) = \{ (x+1, y), (x-1, y), (x, y-1), (x, y+1) \}
\end{equation}
This function takes in a square and returns its neighbors. 
\end{definition}

\begin{theorem}
\label{finitestability}
If the initial board has a finite number of chips, there exists an $N$ and $M$ such that after $N$ big earthquakes or $M$ small earthquakes the resultant board will be stable.
\end{theorem}

We will first show that finite boards cannot return to a previous state. We will then show that for a given finite board there is a maximum nunmber of states that a board can occupy. This shows that all finite boards

So a restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ and $M$ such that for all $m \in \mathbb{Z}$, $BE^N(f) = BE^{N+m}(f)$ and for all $k \in \mathbb{Z}$, $SE^{M}(f) = SE^{M+k}(f)$

\begin{lemma}
If $BE(f) \neq f$, then $BE^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
Let $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ be the following function
\[ \Phi(f) = \sum_{x,y} f(x,y)(x+y)^2 \]
Note that this is well-defined since the boards contain a finite number of chips. We will show that 
\[ E(f) \neq f \Rightarrow \Phi(BE(f)) > \Phi(f) \]
This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.

We can evaluate $\Phi(BE(f)) - \Phi(f)$.
\[ \Phi(BE(f)) = \sum_{x,y} BE(f(x,y)) (x+y)^2 \]
\[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (x+y)^2 \]
Suppose location $f(x,y) \geq 4$, then the number of chips in location $(x,y)$ will have decreased after applying $E$. For each decrease in the number of chips atlocation $(x,y)$ of $4$, there is an increase in its four neighbors neighbors by $1$.

Let $A$ be the set of active locations, or locations which $f(x,y) \geq 4$, which is non-empty by assumption. 
So
\[ \Phi(BE(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(x+y)^2 + 2(x+y+1)^2 + 2(x+y-1)^2 \]
\[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
									   +2x^2 & +2y^2 & +4xy & +2x& +2y &+ 2\\
									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} \]
\[ \geq 4 > 0 \] 
\end{proof}

\begin{lemma}
If $SE(f) \neq f$, then $SE^n(f) \neq f$ for all $n > 0$.
\end{lemma}
Our proof of this lemma is essentiall the same as the preceeding one. We construct a potential function and show that if a small earthquake acts nontrivially on a board then the value of the potential function increases.

\begin{proof}
We use the same definition for $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ 
\[ \Phi(f) = \sum_{x,y} f(x,y)(x+y)^2, \]
and evaluate $\Phi(SE(f)) - \Phi(f)$. Suppose $SE$ effected square $(x_0,y_0)$ and its neighbors, then
\begin{align}
\Phi(SE(f)) - \Phi(f) =\sum_{x,y} SE(f(x,y))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
= SE(f(x_0,y_0))(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} SE(f(x,y))(|x|+|y|)^2 \\- f(x_0,y_0)(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} f(x,y)(|x|+|y|)^2 \\
= 4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2
\end{align}
And because at least two of the neighbors of $(x_0, y_0)$ are one unit farther from the origin than $(x_0, y_0)$ and at most two of the neighbors are one unit closer, and $(|x_0|+|y_0| +1)^2 > (|x_0|+|y_0| - 1)^2$, we can see,
\begin{align}
4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2  \\ 
\geq 4(|x_0| + |y_0|)^2 + 2 (|x_0| + |y_0|+1)^2 + 2 (|x_0| + |y_0|-1)^2 = 4 >0
\end{align}
\end{proof}

\begin{lemma}
\label{finiteextension}
Suppose $\sum_{x,y} f(x,y) = S$, and all chips are within the range $[x_l, x_h] \times [y_l, y_h]$. Then $BE^n(f)$ and $SE^n(f)$ only contain chips within the range $[x_l - \log S, x_h + \log S] \times [y_l-\log S, y_h+\log S]$ for all $n$.
\end{lemma}

@@

\begin{proof}
(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$. Then by Lemma~\ref{finiteextension}, we have that all $E^n(f)$ will lie in $R = [x_l - \log S, x_h + \log S] \times [y_l - \log S, y_h + \log S]$.

There exists less than $(x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ possible boards $b \in \mathcal{B}$ that satisfy this. We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\log S)(y_h - y_l + 2\log S)S$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
\end{proof}
